A Plasma Torus Around a Young Low-Mass Star

### 2025-03-31 paper/ms.tex
update

Regarding the upper limit, Extended Data Figure~\ref{fig:sed} shows the SED.  While
AllWISE \cite{Cutri2014} yielded a confident W3 detection
(9.8$\sigma$) consistent with the photometric extrapolation from bluer
bandpasses, the W4 extraction yielded only a marginal indication
(1.7$\sigma$) of detectable flux.  Similar to other CPVs
\cite{Stauffer2017,Bouma2024}, the photometric uncertainties from WISE
W1 and W2 allow at most a $\lesssim$2\% excess at 3-5\,$\mu$m relative
to the stellar photosphere, and a $\lesssim$5\% excess at 10\,$\mu$m
(W3).  To estimate the implied mass bound, we assume a dust
temperature $T_{\rm d}$=1500\,K, typical for dust near the star (see
\cite{Zhan2019} for discussion regarding dust sublimation).  We then
treat emission from the dust and star as Planck functions, and require
$L_{\rm d} < f L_\star$, where the factor $f$ is set by the
photometric precision of WISE and $L_{\rm d}$ is the bolometric dust
luminosity.  Given the reported uncertainties, we numerically find
$f<6\cdot$10$^{-3}$.  From the Stefan-Boltzmann law we can then write
$A_{\rm d} < f (T_\star / T_{\rm d})^4 Q_{\rm em}^{-1} (4\pi
R_\star)^2$, for $A_{\rm d}$ the total emitting surface area of the
dust, and $Q_{\rm em}$ an emission efficiency parameter.  Converting
this constraint to a dust mass requires an assumption regarding the
grain properties.  We assume a grain density $\rho_{\rm
d}$=3\,g\,cm$^{-3}$ typical for silicate grains, a fixed grain size
$a$=1\,$\mu$m, and no self-absorption.  This enables the assumption
that $A_{\rm d} = N \pi a^2$, for $N$ the total number of dust grains.
This in turn yields an upper limit on the dust mass of
\begin{equation}
  M_{\rm dust} \lesssim 4 \cdot 10^{17}\, {\rm g}\ 
  \left( \frac{f}{6\cdot10^{-3}} \right)
  \left( \frac{T_\star}{3000\,{\rm K}} \frac{1500\,{\rm K}}{T_{\rm d}} \right)^4
  \left( \frac{Q_{\rm em}}{1} \right)^{-1}
  \left( \frac{R_\star}{0.4\,R_\odot} \right)^2
  \left( \frac{a}{1\,\mathrm{\mu m}} \right)
  \left( \frac{\rho_{\rm d}}{3\,\mathrm{g\,cm^{-3}}} \right).
\end{equation}

@@

Broadly, these considerations suggest that if dust is responsible for
the broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.  Given
$M_{\rm gas}$$\approx$8$\times$10$^{16}$\,g from
Section~\ref{subsec:gas}, the allowed dust masses imply $M_{\rm
gas}/M_{\rm dust}$ ranges of 1-100.  More careful measurements of this
ratio---in particular by inferring the dust mass through high
precision infrared spectrophotometry---could provide a path for
distinguishing the scenario of a trapped stellar outflow from an
accumulation of externally-sourced material.  While there are several
plausible external sources, feeding through a low-mass disk in
particular cannot be ruled out based on typical disk depletion times
\cite{Haisch2001}.  Observations of infrared excesses and accretion
signatures in low-mass stars tens of millions of years old suggest a
broad lifetime distribution for such disks
\cite{Silverberg2020,Lee2020,Gaidos2022,Pfalzner2024}.

>>>>

Regarding the upper limit, Extended Data Figure~\ref{fig:sed} shows the SED.  While
AllWISE \cite{Cutri2014} yielded a confident W3 detection
(9.8$\sigma$) consistent with the photospheric extrapolation from bluer
bandpasses, the W4 extraction yielded only a marginal indication
(1.7$\sigma$) of detectable flux.  Similar to other CPVs
\cite{Stauffer2017,Bouma2024}, the photometric uncertainties from WISE
W1 and W2 allow at most a $\lesssim$2\% excess at 3-5\,$\mu$m relative
to the stellar photosphere, and a $\lesssim$5\% excess at 10\,$\mu$m
(W3).  To estimate the implied mass bound, we assume a dust
temperature $T_{\rm d}$=1500\,K, typical for dust near the star (see
\cite{Zhan2019} for discussion regarding dust sublimation).  We then
treat emission from the dust and star as Planck functions, and require
$L_{\rm d} < f L_\star$, where the factor $f$ is set by the
photometric precision of WISE and $L_{\rm d}$ is the bolometric dust
luminosity.  Given the reported uncertainties, we numerically find
$f<6\cdot$10$^{-3}$.  From the Stefan-Boltzmann law we can then write
$A_{\rm d} < f (T_\star / T_{\rm d})^4 Q_{\rm em}^{-1} (4\pi
R_\star)^2$, for $A_{\rm d}$ the total emitting surface area of the
dust, and $Q_{\rm em}$ an emission efficiency parameter.  Converting
this constraint to a dust mass requires an assumption regarding the
grain properties.  We assume a grain density $\rho_{\rm
d}$=3\,g\,cm$^{-3}$ typical for silicate grains, a fixed grain size
$a$=1\,$\mu$m, and no self-absorption.  This enables the assumption
that $A_{\rm d} = N \pi a^2$, for $N$ the total number of dust grains.
This in turn yields an upper limit on the dust mass of
\begin{equation}
  M_{\rm dust} \lesssim 4 \cdot 10^{17}\, {\rm g}\ 
  \left( \frac{f}{6\cdot10^{-3}} \right)
  \left( \frac{T_\star}{3000\,{\rm K}} \frac{1500\,{\rm K}}{T_{\rm d}} \right)^4
  \left( \frac{Q_{\rm em}}{1} \right)^{-1}
  \left( \frac{R_\star}{0.4\,R_\odot} \right)^2
  \left( \frac{a}{1\,\mathrm{\mu m}} \right)
  \left( \frac{\rho_{\rm d}}{3\,\mathrm{g\,cm^{-3}}} \right).
\end{equation}

@@

To summarize, if dust is responsible for
the broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.  Given
$M_{\rm gas}$$\approx$8$\times$10$^{16}$\,g from
Section~\ref{subsec:gas}, the allowed dust masses imply $M_{\rm
gas}/M_{\rm dust}$ ranges of 1-100.  More careful measurements of this
ratio---in particular by inferring the dust mass through high
precision infrared spectrophotometry---could provide a path for
distinguishing the scenario of a trapped stellar outflow from an
accumulation of externally-sourced material.  While there are several
plausible external sources, feeding through a low-mass disk in
particular cannot be ruled out based on typical disk depletion times
\cite{Haisch2001}.  Observations of infrared excesses and accretion
signatures in low-mass stars tens of millions of years old suggest a
broad lifetime distribution for such disks
\cite{Silverberg2020,Lee2020,Gaidos2022,Pfalzner2024}.
